\stdpopsim: a community-maintained standard library of population genetic simulations

### 2019-12-16 popSimManu.tex
added methods bit on Analysis, edited title
few changes from Chris

\title{\stdpopsim: a community toolkit for population genetic simulation}

@@

\subsection*{Model QC procedures}
As a consortium we have agreed to a standardized proceedure for model inclusion
into \stdpopsim that allows for rigorous quality control. Imagine Developer A
wants to introduce a new model into \stdpopsim. Developer A implements the
demographic model for the relevant organism along with a series of required
unit tests and documentation. This model is submitted as a Pull Request,
where it is evaluated by a reviewer and then included as `preliminary',
but is not linked to the online documentation nor the command line interface.
Inclusion of a preliminary model triggers a QC issue, and a second developer,
Developer B, then independently reimplements the model from the relevant
primary sources and adds an automatic unit test for equality between the
QC implementation and the preliminary production model. If the two
implementations are equal the original model is included in \stdpopsim.
If not, we move to an arbitration process whereby the A and B first try
to work out the details of what went wrong. If that fails, the original
authors of the published model A is trying to implement must be contacted
to resolve ambiguities. This process can lead to cycles of revision of
the implementation from Developers A and B but allows the project to assure
quality. Indeed we have already had a number of interesting cases of arbitration
to this point. Further details of our QC process can be found in our
\href{https://stdpopsim.readthedocs.io/en/latest/development.html#}{developer documentation}.

\subsection*{Workflow for analysis of simulated data}
Snakemake pipeline for 1 pop models

\subsection*{Workflow for two population analysis}
Our goal for the two population analysis pipeline was to explore two population inference
on some of the multi-population demographic models implemented in \stdpopsim.
The models we examined include the human Gutenkunst model \citep{gutenkunst2009inferring} as
well as the \emph{Drosophila} Li and Stephan model \citep{li2006inferring}. The human
model is a three population model (Africa, Europe, and Asia) including post-divergence
migration and exponential growth (Figure \ref{fig:IM_popn_human}C), whereas the
\emph{Drosophila} model is a two population model (Africa and Europe) with no post-divergence
migration and constant population sizes (Figure \ref{fig:two_popn_fly}).

To conduct inference on these models, we implemented three commonly-used approaches:
\dadi~\citep{gutenkunst2009inferring}, \fastsimcoal2~\citep{excoffier2013robust},
and \smcpp~\citep{terhorst2017robust}. Our aim was not to exhaustively
test the performance of these methods, but rather to explore their behavior
in two different inference scenarios and demonstrate the potential for this resource to
eventually be used for more in-depth benchmarking. As a result, these methods were used
generally with default settings and we did not attempt to optimize their performance or fit
parameter-rich demographic models. As with the single population inference pipeline,
we developed a \texttt{snakemake}~\citep{koster2012snakemake} workflow to manage
our inference pipeline and maintain reproducibility.

For both \dadi and \fastsimcoal2, our general approach was to fit a two population
isolation with migration (IM) model with constant population sizes to the simulated
data from the human and \emph{Drosophila} models. This IM model contains six parameters:
the ancestral population size, the size of population 1 after the split, the size of
population 2 after the split, the divergence time, and two migration rate parameters.
Importantly, this meant that for both the human and \emph{Drosophila} inference, the
fitted model did not match the simulated model (Figure \ref{fig:IM_popn_human}C and Sx).
In the human case, we therefore performed inference solely on the Africa
and Europe populations, meaning that the Asia population functioned as a ghost
population that was ignored by our inference. Our motivation for fitting this simple
IM model was to mimic the typical approach of two population inference on empirical
data, where the user is not aware of the `true' underlying demography and the inference
model is often misspecified. To ground-truth our inference approach, we also conducted
inference on a generic IM model that was identical to the model used for inference.

To perform inference, we began by running simulations under each model, running
three replicates to inspect for inter-replicate variability. For humans, we ran
whole genome simulations (chromosomes 1-22) under the Gutenkunst model \citep{gutenkunst2009inferring},
using the HapMap II recombination map \citep{international2007second} and taking 20
samples each from the Europe and Africa populations. For \emph{Drosophila},
we simulated 3 Mb of chromosome 2R under the Li and Stephan model \citep{li2006inferring},
using the map inferred by \cite{comeron2012many} and taking 50 samples from each of
the Africa and Europe populations. We tried to run whole genome simulations under
this model but after XXXX amount of time... (JRA fill in details here). For the generic
IM simulations, we used the human genome (chromosomes 1-22) and HapMap II recombination
map \citep{international2007second} and took 20 samples per population.

>>>>

\title{\stdpopsim: a community-maintained standard library of population genetic simulations}

@@

\subsection*{Model QC procedures}
As a consortium we have agreed to a standardized proceedure for model inclusion
into \stdpopsim that allows for rigorous quality control. Imagine Developer A
wants to introduce a new model into \stdpopsim. Developer A implements the
demographic model for the relevant organism along with clear documentation
of the model parameters and populations. This model is submitted as a Pull Request,
where it is evaluated by a reviewer and then included as `preliminary',
but is not linked to the online documentation nor the command line interface.
Inclusion of a preliminary model triggers a QC issue, and a second developer,
Developer B, then independently reimplements the model from the relevant
primary sources and adds an automatic unit test for equality between the
QC implementation and the preliminary production model. If the two
implementations are equal the original model is included in \stdpopsim.
If not, we move to an arbitration process whereby the A and B first try
to work out the details of what went wrong. If that fails, the original
authors of the published model A is trying to implement must be contacted
to resolve ambiguities. This process can lead to cycles of revision of
the implementation from Developers A and B but allows the project to assure
quality. Indeed we have already had a number of interesting cases of arbitration
to this point. Further details of our QC process can be found in our
\href{https://stdpopsim.readthedocs.io/en/latest/development.html#}{developer documentation}.

\subsection*{Workflow for analysis of simulated data}
To demonstrate the utility of \stdpopsim we created \texttt{Snakemake}
workflows \citep{koster2012snakemake} that perform demographic inference on
tree sequence output from our package using a few common software packages. 
Our choice of \texttt{Snakemake} allows complete reproducibility of the 
analyses shown and the entire Analysis repository is available from
\url{https://github.com/popgensims/analysis}.

We performed demographic analysis for one and two population tasks.
Our one population task was to infer historical population size over
time ($N(t)$). This was done using three software packages \stairwayplot
which uses site frequency spectrum information only \citep{liu2015exploring},
the SMC based \MSMC~\citep{schiffels2014inferring} which we ran with two different
sample sizes $n\in (2 , 8)$, and \smcpp~\citep{terhorst2017robust} 
which combines information from the site frequency spectrum with 
recombination information as in SMC-based methods. No attempt
was made at trying to optimize the analysis from
any particular software package in present case,
as our goal was not to benchmark performance of methods but
instead show how such benchmarking could be easily done using
the \stdpopsim resource. In this spirit we ran each software package as near
to default parameters as possibile in each case. For \stairwayplot we
set the parameters numRuns=1 and dimFactor=5000. For \smcpp we used the
``estimate" run mode to infer $N(t)$ with all other parameters set
to their default values. For \MSMC we used the ``--fixedRecombination"
option and used the default number of iterations.

For the single population task we ran human, HomSap, simulations
using a variety of models: OutOfAfricaArchaicAdmixture\_5R19, OutOfAfrica\_3G09,
a constant sized generic model, and a two epoch generic model where the
population size instantaneously decreases from $N=10^4$ to $N=10^3$ 500 
generations before the present. Each HomSap simulation was run
using the HapmapII\_GRCh37 genetic maps. For \emph{D. melanogaster}
we estimated $N(t)$ from an African sample simulated under the DroMel, 
African3Epoch\_1S16 model using the Comeron2012\_dm6 map. Finally we ran
simulations of \emph{A. thaliana} genomes using the AraTha African2Epoch\_1H18
model under the Salome2012\_TAIR7. For each model, 3 replicate whole genomes were
simulated and the population size estimated from those data. In all cases we
set the sample size of the focal population to $N=50$ chromosomes.

Following simulation, low recombination portions of chromosomes were masked
from the analysis in a manner that reflects the ``accessible" subset of sites
used in empirical population genomic studies \citep[e.g.,][]{danecek20111000,langley2012genomic}.
Specifically we masked regions of 1cM or greater in the lowest 5th percentile of the empirical
distribution of recombination, regions which are nearly uniformly absent for
empirical analysis.

Our goal for the two population analysis pipeline was to explore two population inference
on some of the multi-population demographic models implemented in \stdpopsim.
For HomSap we examined the OutOfAfrica_3G09 model with the HapmapII\_GRCh37 genetic map,
and for DroMel we simulated under the OutOfAfrica_2L06 model with the Comeron2012\_dm6.
The HomSap model is a three population model (Africa, Europe, and Asia) including post-divergence
migration and exponential growth (Figure \ref{fig:IM_popn_human}C), whereas the
DroMel model is a two population model (Africa and Europe) with no post-divergence
migration and constant population sizes (Figure \ref{fig:two_popn_fly}).

To conduct inference on these models, we used three commonly used methods:
\dadi~\citep{gutenkunst2009inferring}, \fastsimcoal2~\citep{excoffier2013robust},
and \smcpp~\citep{terhorst2017robust}. As above, these methods were used
generally with default settings and we did not attempt to optimize their performance or fit
parameter-rich demographic models. As with the single population inference pipeline,
we developed a \texttt{snakemake}~\citep{koster2012snakemake} workflow to manage
our inference pipeline and maintain reproducibility.

For both \dadi and \fastsimcoal2, we fit a two population
isolation with migration (IM) model with constant population sizes to the simulated
data from the HomSap and DroMel models. This IM model contains six parameters:
the ancestral population size, the size of population 1 after the split, the size of
population 2 after the split, the divergence time, and two migration rate parameters.
Importantly, this meant that for both species, the
fitted model did not match the simulated model (Figure \ref{fig:IM_popn_human}C and Sx).
In the HomSap case, we therefore performed inference solely on the Africa
and Europe populations, meaning that the Asia population functioned as a ghost
population that was ignored by our inference. Our motivation for fitting this simple
IM model was to mimic the typical approach of two population inference on empirical
data, where the user is not aware of the `true' underlying demography and the inference
model is often misspecified. To ground-truth our inference approach, we also conducted
inference on a generic IM model that was identical to the model used for inference.

From HomSap we ran replicate whole genome simulations under the model
and specified taking 20 samples each from the Europe and Africa populations. 
For DroMel, runtimes under OutOfAfrica_2L06 were prohibitively slow for
running whole genome simulations. Instead we simulated 3 Mb of chromosome 2R.
and take 50 samples from each of
the Africa and Europe populations. We tried to run whole genome simulations under
this model but after XXXX amount of time... (JRA fill in details here). For the generic
IM simulations, we used the HomSap genome along with the HapmapII\_GRCh37 genetic map,
and sampled 20 individuals from each population.
